Forecasting interrupted time series

### 2024-01-08 fits.qmd
Improved naming of methods

---
title: "Forecasting interrupted time series"
author:
- familyname: Rostami-Tabar
  othernames: Bahman
  address:
    - Cardiff Business School
    - Wales CF10 3EU
    - United Kingdom
  email: rostami-tabarb@cardiff.ac.uk
  correspondingauthor: true
- familyname: Hyndman
  othernames: Rob J
  address:
    - Department of Econometrics & Business Statistics
    - Monash University
    - Clayton VIC 3800
    - Australia
  email: Rob.Hyndman@monash.edu
abstract: "A brief summary of our ideas"
keywords: "blah, blah"
wpnumber: no/yr
jelcodes: C10,C14,C22
blind: false
cover: false
toc: false
bibliography: references.bib
biblio-style: authoryear-comp
format:
  wp-pdf:
    include-in-header: header.tex
cite-method: biblatex
number-sections: true
keep-tex: true
fig-height: 5
fig-width: 8
execute:
  echo: false
  warning: false
  message: false
  cache: false
---

@@

## Treat the interruption period as missing

@@

```{r}
#| label: fig-tsol1-plot
#| fig-height: 8
#| fig-pos: "t"
#| fig-cap: "Solution 1. Forecasts from ETS and ARIMA models. Neither works particularly well for disruptions of this magnitude."
tar_read(tsol1_plot)
```

@@

```{r}
#| label: fig-tsol2-plot
#| fig-height: 8
#| fig-pos: "b"
#| fig-cap: Solution 2. We use a level shift from March 2020 to October 2022, and a ramp from October 2021 to October 2022. After October 2022, all intervention variables are set to zero.
tar_read(tsol2_plot)
```

@@

```{r}
#| label: fig-tsol3-plot
#| fig-height: 8
#| fig-pos: "b"
#| fig-cap: Solution 3. The period from March 2020 to October 2022 is set to missing. So the first three years of forecasts show what might have been. The fourth set of forecasts uses observations from November and December 2022, and so the forecasts have been adjusted downwards.
tar_read(tsol3_plot)
```

@@

```{r}
#| label: fig-tsol4-plot
#| fig-height: 8
#| fig-pos: "!b"
#| fig-cap: Solution 4. We replace the observations from March 2020 to October 2022 with the average of the same month in the three years prior to March 2020.
tar_read(tsol4_plot)
```

@@

```{r}
#| label: fig-tensemble-plot
#| fig-height: 8
#| fig-pos: "t"
#| fig-cap: Ensemble solution, combining the previous four approaches shown in Figures [-@fig-tsol1-plot]--[-@fig-tsol4-plot].
tar_read(tensemble_plot)
```

@@

```{r}
#| label: fig-walkers-plot1
#| fig-height: 3
#| fig-cap: "Solution 1: the forecasts do not take into account the lockdown periods, but the ARIMA model is largely able to adjust to the changing level of the series."
tar_read(walkers_plot1)
```

@@

```{r}
#| label: fig-walkers-plot2
#| fig-height: 3
#| fig-cap: "Solution 2: the forecasts are based on an ARIMA model with a dummy variable indicating when the lockdown periods occurred. Many people were in self-imposed lockdown before the start of the first lockdown period, and were reluctant to return to the city after it ended, resulting in poor forecasts on either side of the first lockdown. The model was able to adjust to the changing level of the series during subsequent lockdowns."
tar_read(walkers_plot2)
```

```{r}
#| label: fig-walkers-plot3
#| fig-height: 3
#| fig-cap: "Solution 3: the lockdown periods are set to missing, and the ARIMA model uses the remaining data to produce forecasts. Thus, the forecasts are based on only previous non-lockdown data. Naturally, the forecasts during the lockdown periods do not reflect reality, but they can be interpreted as what might have been."
tar_read(walkers_plot3)
```

```{r}
#| label: fig-walkers-plot4
#| fig-height: 3
#| fig-cap: "Solution 4: the lockdown periods are replaced with estimated counts based on an ARIMA model applied to the remaining data. Then a new ARIMA model is fitted to the whole data set, and used to produce forecasts. As a result, the forecasts during the lockdown periods are not true forecasts, as they use non-lockdown data from the future."
tar_read(walkers_plot4)
```

```{r}
#| label: fig-walkers-ensemble-plot
#| fig-height: 3
#| fig-cap: "Ensemble forecasts based on the first two solutions shown in Figures [-@fig-walkers-plot1] and [-@fig-walkers-plot2]. Other than having some difficulty around the period of the first lockdown, the results are relatively good."
tar_read(walkers_ensemble_plot)
```

>>>>

---
title: "Forecasting interrupted time series"
author:
- familyname: Hyndman
  othernames: Rob J
  address:
    - Department of Econometrics & Business Statistics
    - Monash University
    - Clayton VIC 3800
    - Australia
  email: Rob.Hyndman@monash.edu
  correspondingauthor: true
- familyname: Rostami-Tabar
  othernames: Bahman
  address:
    - Cardiff Business School
    - Wales CF10 3EU
    - United Kingdom
  email: rostami-tabarb@cardiff.ac.uk
abstract: "A brief summary of our ideas"
keywords: "blah, blah"
wpnumber: no/yr
jelcodes: C10,C14,C22
blind: false
cover: false
toc: false
bibliography: references.bib
biblio-style: authoryear-comp
format:
  wp-pdf:
    include-in-header: header.tex
cite-method: biblatex
number-sections: true
keep-tex: true
fig-height: 5
fig-width: 8
execute:
  echo: false
  warning: false
  message: false
  cache: false
---

@@

## Set to missing

@@

```{r}
#| label: fig-tsol1-plot
#| fig-height: 8
#| fig-pos: "!b"
#| fig-cap: "**Highly adaptive models**. Forecasts from ETS and ARIMA models. Neither works particularly well for disruptions of this magnitude."
tar_read(tsol1_plot)
```

@@

```{r}
#| label: fig-tsol2-plot
#| fig-height: 8
#| fig-pos: "!b"
#| fig-cap: "**Intervention model**. We use a level shift from March 2020 to October 2022, and a ramp from October 2021 to October 2022. After October 2022, all intervention variables are set to zero."
tar_read(tsol2_plot)
```

@@

```{r}
#| label: fig-tsol3-plot
#| fig-height: 8
#| fig-pos: "!b"
#| fig-cap: "**Set to missing**. The period from March 2020 to October 2022 is set to missing. So the first three years of forecasts show what might have been. The fourth set of forecasts uses observations from November and December 2022, and so the forecasts have been adjusted downwards."
tar_read(tsol3_plot)
```

@@

```{r}
#| label: fig-tsol4-plot
#| fig-height: 8
#| fig-pos: "!b"
#| fig-cap: "**Estimate what might have been**. We replace the observations from March 2020 to October 2022 with the average of the same month in the three years prior to March 2020."
tar_read(tsol4_plot)
```

@@

```{r}
#| label: fig-tensemble-plot
#| fig-height: 8
#| fig-pos: "!t"
#| fig-cap: "**Ensemble approach**, combining the previous four approaches shown in Figures [-@fig-tsol1-plot]--[-@fig-tsol4-plot]."
tar_read(tensemble_plot)
```

@@

```{r}
#| label: fig-walkers-plot1
#| fig-height: 3
#| fig-cap: "**Highly adaptive model**: the forecasts do not take into account the lockdown periods, but the ARIMA model is largely able to adjust to the changing level of the series."
tar_read(walkers_plot1)
```

@@

```{r}
#| label: fig-walkers-plot2
#| fig-height: 3
#| fig-cap: "**Intervention model**: the forecasts are based on an ARIMA model with a dummy variable indicating when the lockdown periods occurred. Many people were in self-imposed lockdown before the start of the first lockdown period, and were reluctant to return to the city after it ended, resulting in poor forecasts on either side of the first lockdown. The model was able to adjust to the changing level of the series during subsequent lockdowns."
tar_read(walkers_plot2)
```

```{r}
#| label: fig-walkers-plot3
#| fig-height: 3
#| fig-cap: "**Set to missing**: the lockdown periods are set to missing, and the ARIMA model uses the remaining data to produce forecasts. Thus, the forecasts are based on only previous non-lockdown data. Naturally, the forecasts during the lockdown periods do not reflect reality, but they can be interpreted as what might have been."
tar_read(walkers_plot3)
```

```{r}
#| label: fig-walkers-plot4
#| fig-height: 3
#| fig-cap: "**Estimate what might have been**: the lockdown periods are replaced with estimated counts based on an ARIMA model applied to the remaining data. Then a new ARIMA model is fitted to the whole data set, and used to produce forecasts. As a result, the forecasts during the lockdown periods are not true forecasts, as they use non-lockdown data from the future."
tar_read(walkers_plot4)
```

```{r}
#| label: fig-walkers-ensemble-plot
#| fig-height: 3
#| fig-cap: "**Ensemble forecasts** based on the first two solutions shown in Figures [-@fig-walkers-plot1] and [-@fig-walkers-plot2]. Other than having some difficulty around the period of the first lockdown, the results are relatively good."
tar_read(walkers_ensemble_plot)
```

### 2024-01-08 fits.tex
Improved naming of methods

\author{Bahman~Rostami-Tabar, Rob J~Hyndman}
\addresses{\textbf{Bahman Rostami-Tabar}\newline
Cardiff Business School\newline
Wales CF10 3EU\newline
United Kingdom\newline
{Email: rostami-tabarb@cardiff.ac.uk}\newline Corresponding author\newline\\[0.5cm]
\textbf{Rob J Hyndman}\newline
Department of Econometrics \& Business Statistics\newline
Monash University\newline
Clayton VIC 3800\newline
Australia\newline
{Email: Rob.Hyndman@monash.edu}\\[0.5cm]
}

\date{\sf\Date~\Month~\Year}
\makeatletter
 \lfoot{\sf Rostami-Tabar, Hyndman: \@date}
\makeatother

@@

\subsection{Treat the interruption period as
missing}\label{treat-the-interruption-period-as-missing}

@@

\begin{figure}[t]

@@

\caption{\label{fig-tsol1-plot}Solution 1. Forecasts from ETS and ARIMA
models. Neither works particularly well for disruptions of this
magnitude.}

@@

\begin{figure}[b]

@@

\caption{\label{fig-tsol2-plot}Solution 2. We use a level shift from
March 2020 to October 2022, and a ramp from October 2021 to October
2022. After October 2022, all intervention variables are set to zero.}

@@

\caption{\label{fig-tsol3-plot}Solution 3. The period from March 2020 to
October 2022 is set to missing. So the first three years of forecasts
show what might have been. The fourth set of forecasts uses observations
from November and December 2022, and so the forecasts have been adjusted
downwards.}

@@

\caption{\label{fig-tsol4-plot}Solution 4. We replace the observations
from March 2020 to October 2022 with the average of the same month in
the three years prior to March 2020.}

@@

\begin{figure}[t]

@@

\caption{\label{fig-tensemble-plot}Ensemble solution, combining the
previous four approaches shown in Figures
\ref{fig-tsol1-plot}--\ref{fig-tsol4-plot}.}

@@

\caption{\label{fig-walkers-plot1}Solution 1: the forecasts do not take
into account the lockdown periods, but the ARIMA model is largely able
to adjust to the changing level of the series.}

@@

\caption{\label{fig-walkers-plot2}Solution 2: the forecasts are based on
an ARIMA model with a dummy variable indicating when the lockdown
periods occurred. Many people were in self-imposed lockdown before the
start of the first lockdown period, and were reluctant to return to the
city after it ended, resulting in poor forecasts on either side of the
first lockdown. The model was able to adjust to the changing level of
the series during subsequent lockdowns.}

@@

\caption{\label{fig-walkers-plot3}Solution 3: the lockdown periods are
set to missing, and the ARIMA model uses the remaining data to produce
forecasts. Thus, the forecasts are based on only previous non-lockdown
data. Naturally, the forecasts during the lockdown periods do not
reflect reality, but they can be interpreted as what might have been.}

@@

\caption{\label{fig-walkers-plot4}Solution 4: the lockdown periods are
replaced with estimated counts based on an ARIMA model applied to the
remaining data. Then a new ARIMA model is fitted to the whole data set,
and used to produce forecasts. As a result, the forecasts during the
lockdown periods are not true forecasts, as they use non-lockdown data
from the future.}

@@

\caption{\label{fig-walkers-ensemble-plot}Ensemble forecasts based on
the first two solutions shown in Figures \ref{fig-walkers-plot1} and
\ref{fig-walkers-plot2}. Other than having some difficulty around the
period of the first lockdown, the results are relatively good.}

@@

We have not discussed probabilistic forecasting in any detail in this
paper, other than producing some prediction intervals for some of the
examples. One would expect forecast uncertainty to increase during, and
to some extent after, a major disruption. The highly adaptive model and
innovation model should produce relatively good prediction intervals,
although the variance is often underestimated. Because the missing value
approach is producing counterfactual (``what might have been'')
forecasts, the prediction intervals will often not include the actual
observations during the disruption period. Prediction intervals for
ensemble forecasts can be obtained by averaging the end points of the
component prediction intervals, for those models where the prediction
intervals seem reasonable.

>>>>

\author{Rob J~Hyndman, Bahman~Rostami-Tabar}
\addresses{\textbf{Rob J Hyndman}\newline
Department of Econometrics \& Business Statistics\newline
Monash University\newline
Clayton VIC 3800\newline
Australia\newline
{Email: Rob.Hyndman@monash.edu}\newline Corresponding author\newline\\[0.5cm]
\textbf{Bahman Rostami-Tabar}\newline
Cardiff Business School\newline
Wales CF10 3EU\newline
United Kingdom\newline
{Email: rostami-tabarb@cardiff.ac.uk}\\[0.5cm]
}

\date{\sf\Date~\Month~\Year}
\makeatletter
 \lfoot{\sf Hyndman, Rostami-Tabar: \@date}
\makeatother

@@

\subsection{Set to missing}\label{set-to-missing}

@@

@@

\caption{\label{fig-tsol1-plot}\textbf{Highly adaptive models}.
Forecasts from ETS and ARIMA models. Neither works particularly well for
disruptions of this magnitude.}

@@

@@

\caption{\label{fig-tsol2-plot}\textbf{Intervention model}. We use a
level shift from March 2020 to October 2022, and a ramp from October
2021 to October 2022. After October 2022, all intervention variables are
set to zero.}

@@

\caption{\label{fig-tsol3-plot}\textbf{Set to missing}. The period from
March 2020 to October 2022 is set to missing. So the first three years
of forecasts show what might have been. The fourth set of forecasts uses
observations from November and December 2022, and so the forecasts have
been adjusted downwards.}

@@

\caption{\label{fig-tsol4-plot}\textbf{Estimate what might have been}.
We replace the observations from March 2020 to October 2022 with the
average of the same month in the three years prior to March 2020.}

@@

\begin{figure}[!t]

@@

\caption{\label{fig-tensemble-plot}\textbf{Ensemble approach}, combining
the previous four approaches shown in Figures
\ref{fig-tsol1-plot}--\ref{fig-tsol4-plot}.}

@@

\caption{\label{fig-walkers-plot1}\textbf{Highly adaptive model}: the
forecasts do not take into account the lockdown periods, but the ARIMA
model is largely able to adjust to the changing level of the series.}

@@

\caption{\label{fig-walkers-plot2}\textbf{Intervention model}: the
forecasts are based on an ARIMA model with a dummy variable indicating
when the lockdown periods occurred. Many people were in self-imposed
lockdown before the start of the first lockdown period, and were
reluctant to return to the city after it ended, resulting in poor
forecasts on either side of the first lockdown. The model was able to
adjust to the changing level of the series during subsequent lockdowns.}

@@

\caption{\label{fig-walkers-plot3}\textbf{Set to missing}: the lockdown
periods are set to missing, and the ARIMA model uses the remaining data
to produce forecasts. Thus, the forecasts are based on only previous
non-lockdown data. Naturally, the forecasts during the lockdown periods
do not reflect reality, but they can be interpreted as what might have
been.}

@@

\caption{\label{fig-walkers-plot4}\textbf{Estimate what might have
been}: the lockdown periods are replaced with estimated counts based on
an ARIMA model applied to the remaining data. Then a new ARIMA model is
fitted to the whole data set, and used to produce forecasts. As a
result, the forecasts during the lockdown periods are not true
forecasts, as they use non-lockdown data from the future.}

@@

\caption{\label{fig-walkers-ensemble-plot}\textbf{Ensemble forecasts}
based on the first two solutions shown in Figures
\ref{fig-walkers-plot1} and \ref{fig-walkers-plot2}. Other than having
some difficulty around the period of the first lockdown, the results are
relatively good.}

@@

We have not discussed probabilistic forecasting in any detail in this
paper, other than producing some prediction intervals for some of the
examples. One would expect forecast uncertainty to increase during, and
to some extent after, a major disruption. The highly adaptive model and
innovation model should produce relatively good prediction intervals,
although the variance is often underestimated. Because the missing value
approach is producing counterfactual (``what might have been'')
forecasts, the prediction intervals will often not include the actual
observations during the disruption period. Prediction intervals for
ensemble forecasts can be obtained by averaging the end points of the
component prediction intervals \autocite{lichtendahl2013better}, for
those models where the prediction intervals seem reasonable.
